---
title: "Memory and Transporting Objects Analysis"
author: "Clark, Center for Interdisciplinary Statistical Education and Research"
date: "June 26, 2019"
output:
  pdf_document: default
  html_document: default
---
<!-- This is after making updates from the article review -->
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, cache = T, warning = F, message = F)

library(readxl)
library(dplyr)
library(GLMMadaptive)
library(ggplot2)
library(lme4)
library(tidyr)
library(stringr)
library(knitr)
library(forcats)
library(rms)
library(multcomp)
library(foreign)
```

# Subjective Chisquare test

```{r}
subj_reports <- read.spss('./data/Chi Square for Subjective Reports.sav') %>% 
  as.data.frame %>%
  filter(!is.na(Reason))
res <- chisq.test(subj_reports$CupChoice, subj_reports$Reason, correct = F)
summary(res)
table(subj_reports$CupChoice, subj_reports$Reason)
fisher.test( subj_reports$CupChoice == "Procrastinate", subj_reports$Reason == "Efficiency")

```

# Prepare Data

We read in the the memory span data from the file Span Scores and P(Close) Scores.xlsx. The ospan, rotspan and sspan data are scaled by subtracting the mean and dividing by the standard deviation. Furthermore, we compute an average memory score by averaging ospan, rotspan and sspan. 
```{r import}
mem <- read_excel("./data/Span Scores and P(Close) Scores.xlsx")
names(mem) <- c("subject", "p_close", "ospan", "rotspan", "sspan")
mem <- mem[-1,]

mem <- mem %>% 
  mutate_at(vars(ospan, rotspan, sspan), funs(as.numeric(.))) %>%
  # mutate_at(vars(ospan, rotspan, sspan), funs(as.numeric(scale(.)))) %>%
  mutate(
    mem_tot = (ospan + rotspan + sspan),
    subject = subject
  )
```

We read in the MTO logit data from the file Logit_Data.xlsx and the first trial data from the file First Trial PC Frequency. The first trial is joined into the MTO logit data. Update: This is no longer necessary as the Logit_Data was modified to include trial number. 
```{r}
cup <- read_excel("./data/Logit_Data.xlsx")
names(cup) <- c("subject","trial", "dist", "ratio", "load", "choice", "digit")


cup <- cup %>%
  dplyr::select(-load) %>%
  mutate(
    subject = as.character(subject)
    ) %>%
  left_join(
  mem %>% dplyr::select(subject, ospan, rotspan, sspan, mem_tot)
)

ratio_translation <- c("50:100" = 0.5, "100:100" = 1, "100:50" = 2)

ftry <- read_excel("./data/First Trial PC Frequency.xlsx")[1:12,]
ix <- sapply(ftry[,3:(ncol(ftry)-3)], function(x) which(!is.na(x))) %>% as.numeric

ftry <- data.frame(subject = colnames(ftry[,3:(ncol(ftry)-3)]), 
                   dist = ftry$Distance[ix], ratio = ftry$Ratio[ix])
ftry <- ftry %>%
  mutate(
    dist = str_replace(dist, "'", ""),
    subject = as.character(subject),
    first = 1L,
    ratio = as.character(ratio),
    ratio = as.numeric(ratio_translation[ratio])
  )

cup <- cup %>%
  left_join(
    ftry
  ) %>%
  mutate(
    first = if_else(is.na(first), 0L, first, missing = 0L),
    ratio = factor(ratio)
  )


```

# Correlations

We compute pearson correlations between each pair of the 3 memory tasks. The correlations and p-values are shown in the output below. 
```{r}
cor.test(mem$ospan, mem$rotspan)
cor.test(mem$ospan, mem$sspan)
cor.test(mem$rotspan, mem$sspan)
```

# Mixed effects model with the average of the memory tasks

We fit a logistic mixed effects model to relate the choice of cup to the mem_tot, ratio and interaction between mem_tot:ratio. We conduct likelihood ratio tests for an overall effect of mem_tot, an overall effect of ratio and an interaction between mem_tot and ratio. Below is the model summary along with a table giving the likelihood ratio test statistic along with the p-value for each effect. Note that the mem_tot and ratio rows in the table are testing both the main effect and interaction.

```{r}
scaling <- list(mem_tot_mean = mean(cup$mem_tot), mem_tot_sd = sd(cup$mem_tot))
cup <- cup %>%
  mutate(
    choice_far = as.integer(choice == 0),
    mem_tots = (mem_tot - mean(mem_tot))/sd(mem_tot))
mod <- mixed_model(choice ~  mem_tots + ratio + mem_tots:ratio, random = ~ 1|subject, 
             family = binomial(link = "logit"), data = cup)
# mod0 <- glmer(choice ~  mem_tots + ratio + mem_tots:ratio + (1|subject), 
#              family = binomial(link = "logit"), data = cup)


terms <- c("mem_tots", "ratio", "mem_tots:ratio")
comp_mod <- function(terms, mod, ...) {
  form <- as.formula(paste0("choice ~ ", paste(terms, collapse = "+")))
  # mod_i <- glmer(form, family = binomial(link = "logit"), data = cup, ...)
  mod_i <- mixed_model(form, random = ~ 1|subject, family = binomial(link = "logit"), data = cup, ...)
  an <- anova(mod_i, mod)
  c(an$LRT, an$p.value)
}

summary(mod)
K <- rbind(
  "ratio.5" = c(1, 0, 0, 0, 0, 0),
  "slope.5" = c(0, 1, 0, 0, 0, 0),
  "ratio1" = c(1, 0, 1, 0, 0, 0),
  "slope1" = c(0, 1, 0, 0, 1, 0),
  "ratio2" = c(1, 0, 0, 1, 0, 0),
  "slope2" = c(0, 1, 0, 0, 0, 1),
  "ratio1-.5" = c(0, 0, 1, 0, 0, 0),
  "ratio2-.5" = c(0, 0, 0, 1, 0, 0),
  "ratio2-1"  = c(0, 0, -1, 1, 0, 0)
)
wrap <- function(obj) {
  lst <- list(obj)
  attr(lst, "class") <- "wrap"
  lst
}
vcov.wrap <- function(obj,...) {
  vc <- vcov(obj[[1]])
  n <- dim(vc)[1]
  vc[-n, -n]
}
coef.wrap <- function(obj, ...) {
  coef(summary(obj[[1]]))[,1]
}
model.matrix.wrap <- function(obj, ...) {
  model.matrix(obj[[1]])
}
model.frame.wrap <- function(obj, ...) {
  model.frame(obj[[1]])
}
terms.wrap <- function(obj, ...) {
  terms(obj[[1]])
}
w <- wrap(mod)
(sm <- summary(glht(w, K), test = univariate()))
round(binomial()$linkinv(sm$test$coefficients[c(1,3,5)]),3)
round(exp(sm$test$coefficients[c(2,4,6)]),3)
cbind(sm$test$tstat^2, sm$test$pvalues)

1/exp(confint(mod))
```

## ANOVA Table
```{r}
aov_table <- cbind(
  mem_tots = comp_mod(terms[-c(1, 3)], mod),
  ratio = comp_mod(terms[-c(2, 3)], mod),
  interaction = comp_mod(terms[-c(3)], mod)
)
row.names(aov_table) <- c("LRT", "p-value")

t(round(aov_table, 3))
```


# Plot of memory span effect
We use the model with mem_tots and ratio (described above) to plot the estimated average log odds of choosing the first cup first for the 3 different ratios, and memory span scores ranging between 0 and 1. 

```{r}
mc <- marginal_coefs(mod, std_errors = T)
avg_lo <- function(mem_tot) {
  mem_tots <- (mem_tot - scaling$mem_tot_mean) / scaling$mem_tot_sd
  contr1 <- rbind("50:100" = c(1, mem_tots, 0, 0, 0, 0),
                  "100:100" = c(1, mem_tots, 1, 0, mem_tots, 0),
                  "100:50" = c(1, mem_tots, 0, 1, 0, mem_tots))
  beta <- as.matrix(mc$betas)
  est <- contr1 %*% beta  
  vc <- mc$var_betas
  vars <- contr1 %*% vc %*% t(contr1) %>% diag
  # p-values for (0.5 group, 1.0 group, 2.0 group)
  pvals <- 1 - pnorm(est / sqrt(vars))
  # pvals <- pmin(pnorm(est / sqrt(vars)), 1 - pnorm(est / sqrt(vars))) * 2
  
  # upper ci for (0.5 group, 1.0 group, 2.0 group)
  high <- est + qnorm(0.975)*sqrt(vars)
  # lower ci for (0.5 group, 1.0 group, 2.0 group)
  low <- est - qnorm(0.975)*sqrt(vars)
  
  df <- data.frame(est = est, high = high, low = low)
  df$ratio <- row.names(df)
  df
}


mem_totv <- 28:63 #seq(-2,2, by = 1) # 28:63
res <- do.call(rbind,lapply(mem_totv, avg_lo))
res$mem_totv <- rep(mem_totv, each = 3)

res %>%   
  mutate(
    ratio = fct_relevel(ratio, c("50:100", "100:100", "100:50")),
    prob = binomial()$linkinv(est),
    low = binomial()$linkinv(low),
    high = binomial()$linkinv(high)
  ) %>%
  ggplot(aes(x = mem_totv, y = prob, ymin = low, ymax = high, colour = ratio)) +  geom_line() + geom_ribbon(alpha = 0.3) +
  labs(y = "Probability of Selecting Close Cup First", x = "Working Memory Span") + labs(colour = "ratio") + 
  scale_y_continuous(labels = c("0%", "25%", "50%", "75%", "100%"),limits = c(0,1)) + facet_wrap(~ratio) + theme(legend.position="none")

```

# Two sided test for precrastination
We use the model with mem_tots and ratio (described above) to conduct Wald tests to test whether the average log odds of choosing the first cup first is equal to zero. We also conduct this test separately for each ratio. 

```{r}

contr1 <- rbind("combined" = c(1, 0, 1/3, 1/3, 0, 0),
                "50:100" = c(1, 0, 0, 0, 0, 0),
                "100:100" = c(1, 0, 1, 0, 0, 0),
                 #"100:50" = c(0, 1, 0, 0, 1, 0))
                "100:50" = c(1, 0, 0, 1, 0, 0))

beta <- as.matrix(mc$betas)#beta <- matrix(coefficients(summary(mod))[,1])
est <- contr1 %*% beta
vc <- mc$var_betas
vars <- contr1 %*% vc %*% t(contr1) %>% diag
# p-values for (0.5 group, 1.0 group, 2.0 group)
(waldstat <- est / sqrt(vars))
pvals <- 2*(1 - pnorm(abs(waldstat)))
# pvals <- pmin(pnorm(est / sqrt(vars)), 1 - pnorm(est / sqrt(vars))) * 2

# upper ci for (0.5 group, 1.0 group, 2.0 group)
high <- est + qnorm(0.975)*sqrt(vars)
# lower ci for (0.5 group, 1.0 group, 2.0 group)
low <- est - qnorm(0.975)*sqrt(vars)

data.frame(pvals, high, low) %>% 
  mutate(prob = 1 / (1 + exp(-est)),
         plow = 1 / (1 + exp(-low)),
         phigh = 1 / (1 + exp(-high))) %>%
  kable(digits = 5)

```

# Post-hoc tests for ratio

We use the model with mem_tots and ratio (described above) the method of Westfall 1997 to conduct post-hoc tests between all 3 ratios.
```{r}
library(multcomp)
summary(glht(w, mcp(ratio = "Tukey")), test = adjusted(type = "Westfall"))
```


# Model with spline for trial number

We fit a logistic mixed effects model to relate the choice of cup to mem_tots, ratio and trial number. The trial # is included as a natural cubic spline with three knots. All second order interactions are included, however, the interactions with trial are limited to being linear in the effect of trial#. We conduct a likelihood ratio test ANOVA to test for an overall effect of each variable along with testing each of the three interactions.

```{r}
mycontrol <- glmerControl(optimizer="bobyqa",
                            optCtrl=list(maxfun=2e5))
cup$ratiog <- cup$ratio == "0.5"

terms <- c("mem_tots", "ratio", "mem_tots:ratio", "rcs(trial, 3)", "mem_tots:trial", "ratio:trial")
form1 <- choice ~  mem_tots + ratio + mem_tots:ratio + rcs(trial, 3) + mem_tots:trial + 
  ratio:trial + (1|subject)

mod <- glmer(form1, 
             family = binomial(link = "logit"), data = cup, control=mycontrol)
# mod_rms <- lrm(choice ~  mem_avg + ratio + mem_avg:ratio + rcs(trial, 3) + subject, data = cup %>% mutate(subject = factor(subject)), penalty = 1)
# cov2cor(vcov(mod)) %>% as.matrix %>% image
# cov2cor(vcov(mod)) %>% as.matrix %>% View
summary(mod)

```

## ANOVA Table
```{r}
comp_mod <- function(terms, mod, ...) {
  form <- as.formula(paste0("choice ~ ", paste(terms, collapse = "+"), "+ (1|subject)"))
  mod_i <- glmer(form, family = binomial(link = "logit"), data = cup, ...)
  an <- anova(mod_i, mod)
  c(an$Chisq[2], an$`Pr(>Chisq)`[2])
}

aov_table <- cbind(
  mem_avg = comp_mod(terms[-c(1, 3, 5)], mod),
  ratio = comp_mod(terms[-c(2, 3, 6)], mod, control = mycontrol),
  trial = comp_mod(terms[-c(4:6)], mod, control = mycontrol),
  ratioXmem = comp_mod(terms[-c(3)], mod, control = mycontrol),
  trialXmem = comp_mod(terms[-c(5)], mod, control = mycontrol),
  ratioXtrial = comp_mod(terms[-c(6)], mod, control = mycontrol)
)
row.names(aov_table) <- c("LRT", "p-value")

t(round(aov_table, 3))

```



# Trial Number Plot

```{r}
# First we want to produce confidence intervals on the average of the odds
# Next we want to produce prediction intervals


#1st contr (mem_avg = 0, ratio = 0.5, trial = 1, dist = 0)

#  [1] "(Intercept)"         "mem_avg"             "ratio1"             
#  [4] "ratio2"              "rcs(trial, 3)trial"  "rcs(trial, 3)trial'"
#  [7] "dist12-22"           "dist6-12"            "dist6-16"           
# [10] "mem_avg:ratio1"      "mem_avg:ratio2"      "mem_avg:trial"      
# [13] "trial:ratiogTRUE"

mm <- model.matrix(choice ~  mem_tots * ratio + rcs(trial, 3) + mem_tots:trial + ratio:trial, data = cup)

emgrid <- as.data.frame(mm) %>% dplyr::select(ratio1, ratio2, `rcs(trial, 3)trial`, `rcs(trial, 3)trial'`) %>% unique %>% arrange(`rcs(trial, 3)trial`) %>%
  group_by_all() %>%
  do(data.frame(mem_tots = c(-1, 0, 1))) %>%
  ungroup

emgrid <- with(emgrid, cbind(1, mem_tots, ratio1, ratio2, `rcs(trial, 3)trial`, `rcs(trial, 3)trial'`, mem_tots*ratio1, mem_tots*ratio2, mem_tots*`rcs(trial, 3)trial`, ratio1 * `rcs(trial, 3)trial`, ratio2 * `rcs(trial, 3)trial`))

lodds <- emgrid %*% coef(summary(mod))[,1]
model_avg <- emgrid %>% as.data.frame %>%
  rename(trial = `rcs(trial, 3)trial`) %>%
  mutate(
    ratio = interaction(ratio1, ratio2),
    ratio = fct_recode(ratio,"50:100"="0.0","100:100"="1.0", "100:50"="0.1"),
    lodds = lodds[,1]
  ) %>%
  dplyr::select(trial, ratio, lodds, mem_tots)

vbeta <- emgrid %*% vcov(mod) %*% t(emgrid)
sbeta <- sqrt(diag(vbeta))
model_avg <- model_avg %>%
  mutate(
    sbeta = sbeta,
    odds = exp(lodds),
    prob = odds / (1+odds),
    low = lodds + qnorm(0.025)*sbeta,
    high = lodds + qnorm(0.975)*sbeta
  )

 # mem_tots = fct_recode(mem_tots,paste0("+1 (M=56)"= "1", "0 (M=47)" = "0", "-1 (M=39)" = "-1"))
model_avg %>% 
  mutate(mem_tots = factor(mem_tots, levels = c("-1", "0", "1"))
        ) %>%
  ggplot(aes(trial, y = lodds,  ymin = low, ymax = high)) + geom_point(aes(shape = mem_tots), size = 3) + scale_shape_manual(values=c(1,17,15)) + geom_line(aes(linetype = mem_tots)) + scale_x_continuous(breaks = 1:12, labels = c("1","","","","","6", "", "","","","","12")) + labs(y = "log odds", x = "trial #") + facet_grid(~ratio) + labs(y = "Log Odds of Selecting Close Cup First", x = "Trial Number") + labs(shape = "Working \nMemory Span", linetype = "Working \nMemory Span") +
  scale_y_continuous(limits = c(-5.6,2.6))+theme(text = element_text(size=15))#+ theme(axis.text.x=element_text(angle=45))

```


# References
Peter H. Westfall (1997), Multiple testing of general contrasts using logical constraints and correlations. Journal of the American Statistical Association, 92, 299–306.